---
title: "Mice Protein Expression Analysis"
author: "Yilan Li"
output:
  html_document:
    df_print: paged
    theme: darkly
    toc: yes
    toc_folate: yes
    code_folding: hide
    collapse: true
    fig_width: 6
    fig_height: 4
    locale: en_US.UTF-8
---

```{r,message=FALSE, warning=FALSE}
library(tidyverse)
data <- read.csv("data.csv")
```

## Data Description

This dataset comes from the **UC Machine Learning Repository** [Mice Protein Expression](https://archive.ics.uci.edu/dataset/342/mice+protein+expression). It consists of expression levels of 77 proteins modifications that produced signals in the nuclear fraction of the cortex. The dataset includes measurements from **38 control mice** and **34 trisomic mice** (Down syndrome).

```{r}
data
```

### What are the columns represent

**Mouse ID**: Unique identifier for each mouse, **15 measurements** were taken for each protein per mouse.

**Expression Levels**: The dataset contains values for the expression levels of **77 proteins**.

**Genotype**: The genotype of each mouse, either **control (c)** or **trisomy (t)**.

**Treatment Type**: The treatment administered to the mouse, either **memantine (m)** or **saline (s)**.

**Behavior**: Describes the behavioral context of each mouse, either **context-shock (CS)** or **shock-context (SC)**.

### Classes

The mice are classified into eight groups based on features such as **genotype**, **behavior**, and **treatment**. The genotypes can either be **control** or **trisomic**. In terms of behavior, some mice have been stimulated to learn (context-shock), while others have not (shock-context). Additionally, in order to assess the effect of the drug **memantine** on the ability to learn in trisomic mice, some mice were injected with the drug, and others were injected with saline as control.

```{r, message=FALSE, warning=FALSE}
blue_palette <- colorRampPalette(c("darkblue", "lightblue"))(8)

ggplot(data, aes(class)) +
  geom_bar(aes(fill = class), alpha = 0.8) +
  scale_fill_manual(values = blue_palette) +
  labs(title = "Number of Classes", y = "Count", x="Class")
```

-   **c-CS-s**: Control mice, stimulated to learn, injected with saline (9 mice)

-   **c-CS-m**: Control mice, stimulated to learn, injected with memantine (10 mice)

-   **c-SC-s**: Control mice, not stimulated to learn, injected with saline (9 mice)

-   **c-SC-m**: Control mice, not stimulated to learn, injected with memantine (10 mice)

-   **t-CS-s**: Trisomic mice, stimulated to learn, injected with saline (7 mice)

-   **t-CS-m**: Trisomic mice, stimulated to learn, injected with memantine (9 mice)

-   **t-SC-s**: Trisomic mice, not stimulated to learn, injected with saline (9 mice)

-   **t-SC-m**: Trisomic mice, not stimulated to learn, injected with memantine (9 mice)

## Data Cleaning and handle missing values

The dataset contains some missing values. In this case, we will handle the missing values by replacing them with the mean of the corresponding protein group.

```{r,message=FALSE, warning=FALSE}
missing_values <- data %>%
  setNames(gsub("_N", "", names(.))) %>%
  summarise(across(everything(), ~ sum(is.na(.)), .names = "{.col}")) %>%
  pivot_longer(everything(), names_to = "key", values_to = "missing") %>%
  mutate(
    total = nrow(data),
    pct = (missing / total) * 100,
    isna = missing > 0
  ) %>%
  filter(isna) %>%
  arrange(desc(pct))

percentage.plot <- missing_values %>%
  ggplot(aes(x = reorder(key, pct), y = pct)) +
  geom_bar(stat = 'identity', fill = 'blue', alpha = 1) +
  scale_x_discrete(limits = missing_values$key) +
  coord_flip() +
  labs(title = "Missing Values In the Dataset", x = 'Protein Name', y = "Percentage of Missing Values") +
  theme(plot.title = element_text(hjust = 0.5)) + 
  guides(fill = "none")
percentage.plot
```

```{r,message=FALSE, warning=FALSE}
data <- data %>%
  group_by(class) %>%
  mutate(across(everything(), ~ replace(., is.na(.), mean(., na.rm = TRUE)))) %>%
  as.data.frame()

names(data) <- gsub("_N", "", names(data))
proteins <- names(data[2:78])
classes <- as.vector(unique(as.character(data$class)))
```

## Data Visualization

### Histogram for Expression Levels {.tabset .tabset-fade .tabset-pills}

```{r, results='asis', warning = FALSE, message = FALSE}
for(i in 1:length(proteins)){
  plot <- ggplot(data, aes(eval(parse(text = proteins[i])))) +
    geom_histogram(aes(fill=..count..), color = "black", alpha = 0.5) +
    scale_fill_gradient("Count", low="blue", high="darkblue") +
    labs(title = proteins[i],
         x = "Protein Expression Level",
         y = "Frequency") +
    theme_light()
  cat("#### ", proteins[i], "\n")
  print(plot)
  cat('\n\n')
}
```

### Boxplot for Distribution of Protein Expression Level by Class {.tabset .tabset-fade .tabset-pills}

```{r, results='asis', warning = FALSE, message = FALSE}
for(i in 1:length(proteins)){
  plot <- ggplot(data, aes(x = class, y = eval(parse(text = proteins[i])))) +
    geom_boxplot(aes(fill = class), alpha = 0.8) +
    stat_summary(fun.y=mean, colour="black", geom="point", size=1.5) +
    labs(y = proteins[i]) +
    scale_fill_manual(values = blue_palette)
  cat("#### ", proteins[i], "\n")
  print(plot)
  cat('\n\n')
}
```
Outliers will be removed before performing Anova and tukey's HSD.

### Anova and TukeyHSD Test {.tabset .tabset-fade .tabset-pills}

```{r, results='asis', warning = FALSE, message = FALSE}
library(broom)

alpha <- 0.01

aov_test <- function(prot) {
  aov_result <- aov(eval(parse(text = prot)) ~ class, data)
  tidy(aov_result)
}

tukey_test <- function(aov_result) {
  TukeyHSD(aov_result)$class
}

AOV_df <- data.frame(matrix(NA, nrow = length(proteins), ncol = 3))
names(AOV_df) <- c("Protein", "Var_expl", "p.value")

for (i in 1:length(proteins)) {
  cat("#### ", proteins[i], "\n")
  aov_result <- aov(eval(parse(text = proteins[i])) ~ class, data)
  AOV <- tidy(aov_result)
  
  AOV_df$Protein[i] <- proteins[i]
  AOV_df$Var_expl[i] <- round(AOV$sumsq[1] / (AOV$sumsq[1] + AOV$sumsq[2]), 2)
  AOV_df$p.value[i] <- AOV$p.value[1]
  
  cat(paste0("Total variability:  ", round(AOV$sumsq[1] + AOV$sumsq[2], 2)), "\n")
    cat('\n\n')
  cat(paste0("Percentage of explained variability for ", proteins[i], ":  ", round(AOV$sumsq[1] * 100 / (AOV$sumsq[1] + AOV$sumsq[2]), 2), "%\n"))
    cat('\n\n')
  cat(paste0("P value for ", proteins[i], ":  ", round(AOV$p.value[1], 5)), "\n\n")
    cat('\n\n')
  
  if (AOV$p.value[1] < alpha) {
    cat("Significant ANOVA result; proceeding to Tukey HSD test for ", proteins[i], "\n")
    tukey_result <- tukey_test(aov_result)
    print(tukey_result)
  } else {
    cat("No significant ANOVA result; skipping Tukey HSD test for ", proteins[i], "\n")
  }
  
  cat('\n\n')
}
```


```{r, results='asis', warning = FALSE, message = FALSE}
for (i in 1:length(proteins)) {
  cat("#### ", proteins[i], "\n")
  
  # Perform ANOVA
  aov_result <- aov(eval(parse(text = proteins[i])) ~ class, data = data)
  
  # Output the ANOVA summary
  print(summary(aov_result))
  
  # Perform Tukey's HSD if ANOVA is significant
  if (summary(aov_result)[[1]][["Pr(>F)"]][1] < 0.05) {
    tukey_result <- TukeyHSD(aov_result)
    print(tukey_result)
  }
  
  cat("\n\n")
}
```